---
title: "ORISMA: Mapping Occupational Risk Evidence in Metal Additive Manufacturing"
author: 
  - "PhD. Raúl Aguilar-Elena (GPRL, Universidad Internacional de Valencia)"
  - "Ana Delgado-García (Universidad de Salamanca)"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{ORISMA: Mapping Occupational Risk Evidence in Metal Additive Manufacturing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  warning   = FALSE,
  message   = FALSE,
  eval      = FALSE
)
```

# Introduction

Metal additive manufacturing (AM) -- commonly known as 3D metal printing --
represents one of the most rapidly expanding advanced manufacturing technologies
of the 21st century. Its occupational health implications are still poorly
understood, partly because the scientific literature is growing faster than the
capacity of occupational health professionals to synthesise and apply it.

`orisma` (**O**ccupational **R**isk **I**ntegrated **S**ystematic **M**apping and
**A**nalysis) addresses this gap by automating the complete pipeline from
raw bibliographic exports to actionable risk intelligence -- in a single
function call.

This vignette demonstrates the full ORISMA workflow using a systematic corpus
of 184 records (114 unique after deduplication) retrieved from Web of Science
and Scopus on the topic of occupational health and safety in metal additive
manufacturing (2015-2026).

## What ORISMA does

ORISMA automates seven analytical steps:

1. **Ingestion** -- reads RIS/BibTeX/CSV files from multiple databases
2. **Deduplication** -- three-step pipeline (DOI + title + fuzzy matching)
3. **Risk extraction** -- dictionary-based classification into 58 normative categories
4. **Bibliometric analysis** -- computes WRDI, RCS, and MGP indicators
5. **Dimension detection** -- automatic discovery of normative blocks (A-F)
6. **Preventive intelligence** -- Abstract Sufficiency Score (ASS) and bridge article detection
7. **Report generation** -- bilingual HTML reports, risk sheets, extraction matrices

## Installation

```{r install, eval = FALSE}
# From CRAN
install.packages("orisma")

# Development version
remotes::install_github("Aguilar-Elena/orisma")
```

---

# The three-line workflow

The entire analysis pipeline runs in three lines of R:

```{r three_lines}
library(orisma)

refs   <- orm_load("path/to/ris/files/")
result <- orm_run(refs, topic = "Metal additive manufacturing - OHS")
orm_report(result, lang = "en", out_dir = "my_outputs/")
```

The remainder of this vignette unpacks what happens inside each step and
interprets the results from the metal AM corpus.

---

# Step 1: Loading and deduplication

## Loading references

`orm_load()` reads all RIS, BibTeX, and CSV files from a folder, detects the
source database automatically, and returns a standardised `orisma_refs` object.

```{r load}
library(orisma)

refs <- orm_load(
  path = "path/to/ris/files/",
  lang = "en"
)
```

In the metal AM corpus, two files were loaded:

- `Wos.ris` -- 73 records from Web of Science
- `scopusris.ris` -- 111 records from Scopus
- **Total: 184 records identified**

## Deduplication

`orm_dedup()` runs a three-step deduplication pipeline:

1. **Exact DOI match** -- records with identical DOIs are deduplicated
2. **Normalised title match** -- titles are lowercased, stripped of punctuation, and matched exactly
3. **Fuzzy match** -- remaining potential duplicates are identified using Levenshtein distance with a configurable threshold (default: 0.90 similarity)

```{r dedup}
# Deduplication runs automatically inside orm_run()
# but can be called independently:
deduped <- orm_dedup(refs, fuzzy_threshold = 0.90, verbose = TRUE)
```

**Result for the metal AM corpus:**

| Step | Duplicates removed | Records remaining |
|------|-------------------|-------------------|
| Exact DOI | 70 | 114 |
| Normalised title | 0 | 114 |
| Fuzzy match | 0 | 114 |

**70 duplicates (38.0%)** were removed, yielding **114 unique records**.
The high overlap between WoS and Scopus confirms the importance of
multi-database deduplication in systematic reviews of this topic.

---

# Step 2: Risk category extraction

## The ORISMA dictionary

Risk categories are extracted using a built-in normative dictionary of
**58 categories** organised in six blocks anchored in ISO 45001:2018,
INSST, NIOSH, and EU-OSHA:

```{r dict}
dict <- orm_dict()
orm_dict_categories(dict)
```

| Block | Categories | Normative anchors |
|-------|-----------|-------------------|
| A -- Safety at work | 18 | INSST / ISO 45001 |
| B -- Industrial hygiene | 8 | INSST / NIOSH |
| C -- Ergonomics | 8 | INSST / ISO 45001 |
| D -- Psychosociology | 11 | INSST / ISO 45001 |
| E -- Biological hazards | 5 | EU-OSHA / NIOSH |
| F -- Emerging technologies | 8 | EU-OSHA 2024-2026 |

The dictionary can be extended for any domain:

```{r dict_extend}
# Add terms to an existing category
dict <- orm_dict_add_terms(dict, "nanomaterials",
                           c("nano-aerosol", "NOAA particle"))

# Add a completely new category
dict <- orm_dict_add_category(dict,
  key      = "laser_safety",
  label_en = "Laser safety in AM processes",
  label_es = "Seguridad laser en procesos AM",
  terms    = c("laser safety", "laser hazard", "laser exposure")
)
```

## Extraction results

```{r extract}
mx <- orm_extract(deduped, dict = dict,
                  fields  = c("title", "abstract", "keywords"),
                  verbose = TRUE)
print(mx)
```

**22 of 58 categories** were detected in the metal AM corpus. The top five
by frequency:

| Category | N records | % |
|----------|-----------|---|
| Additive manufacturing and 3D printing | 113 | 99.1% |
| Exposure to hazardous chemical agents | 61 | 53.5% |
| Nanomaterials and nanotechnology | 28 | 24.6% |
| Exposure to carcinogens, mutagens, reprotoxics (CMR) | 15 | 13.2% |
| Exposure to ionising radiation | 10 | 8.8% |

Note that "Additive manufacturing and 3D printing" (Block F) dominates the
corpus -- this is the **technological context** of the corpus, not a risk
category per se. ORISMA's `orm_priority()` function automatically separates
context categories from risk categories.

---

# Step 3: The three original ORISMA indicators

## Worker-Risk Disconnection Index (WRDI)

The WRDI measures the proportion of studies characterising an occupational risk
**without** including direct worker exposure data. It is computed at both the
category level and globally.

$$WRDI_c = 1 - \frac{N_{workers,c}}{N_{total,c}}$$

where $N_{workers,c}$ is the number of studies in category $c$ that include
worker exposure terms (e.g. "worker exposure", "breathing zone", "personal
sampling", "field study") and $N_{total,c}$ is the total number of studies.

```{r wrdi}
result <- orm_run(refs, topic = "Metal additive manufacturing - OHS")
cat("Global WRDI:", result$WRDI_global, "\n")
```

**Global WRDI = 0.6053** for the metal AM corpus. This means that
**60.5% of studies** characterise occupational risks without including
direct worker exposure data. The literature is predominantly technical
-- characterising emissions, particles, and processes in controlled
settings -- without measuring what workers actually breathe or absorb.

WRDI by category reveals important heterogeneity:

| Category | WRDI | Interpretation |
|----------|------|---------------|
| Indoor air quality | 0.22 | Best connected to workers |
| Exposure to ionising radiation | 0.30 | Good worker data |
| Exposure to hazardous chemical agents | 0.34 | Moderate |
| Nanomaterials and nanotechnology | 0.39 | Moderate |
| Biological hazard - Bacteria | 1.00 | No worker data at all |
| Artificial intelligence and autonomous systems | 1.00 | No worker data |

## Risk Category Saturation Index (RCS)

The RCS measures the relative dominance of each risk category compared
to a hypothetical uniform distribution:

$$RCS_c = \frac{N_c \cdot K}{N_{total}}$$

where $K$ = 58 categories. RCS > 1 indicates over-representation;
RCS < 1 indicates under-representation.

```{r rcs}
ind <- result$indicators[result$indicators$n_records > 0, ]
ind[order(-ind$RCS), c("label", "n_records", "RCS")]
```

The metal AM corpus shows extreme saturation:

- **Additive manufacturing and 3D printing: RCS = 24.1** -- massively
  over-represented (the dominant topic of the corpus)
- **Exposure to hazardous chemical agents: RCS = 13.0** -- strongly
  over-represented
- **Most other categories: RCS < 1** -- under-represented or absent

This pattern confirms that the literature has concentrated on chemical
emissions while largely ignoring safety, ergonomic, psychosocial, and
biological risks.

## Material-Gap Profile (MGP)

The MGP identifies materials that are hazardous but understudied:

$$MGP_m = \frac{hazard\_proxy_m}{coverage_m}$$

```{r mgp}
print(result$MGP)
```

| Material | Coverage | Hazard proxy | MGP |
|----------|----------|--------------|-----|
| Titanium | 2.9% | 1.33 | **45.3** |
| Aluminium | 8.8% | 1.11 | 12.6 |
| Nickel/Inconel | 8.8% | 0.89 | 10.1 |
| Steel | 17.6% | 1.33 | 7.6 |
| Others/Mixed | 61.8% | 0.89 | 1.4 |

**Titanium has the highest MGP (45.3)**. Despite its known carcinogenic
and respiratory hazard potential, it appears in only 2.9% of corpus
studies. This represents the most critical material gap in the field.

---

# Step 4: Automatic dimension detection

`orm_autodim()` automatically discovers the normative blocks present in
the corpus without any user configuration. In "blocks" mode (default), it
uses the six normative blocks (A-F) as dimensions:

```{r autodim}
dims <- orm_autodim(result$mx, method = "blocks", verbose = TRUE)
print(dims)
```

For the metal AM corpus, all six blocks are present, with Block F
(Emerging technologies) dominating (n=113) followed by Block B
(Industrial hygiene, n=66).

`orm_dim_matrix()` then builds a risk category x normative block
cross-matrix, visualised as a hierarchical clustered heatmap:

```{r dim_matrix}
mat <- orm_dim_matrix(result, dims,
                      out_dir = "outputs/plots/",
                      lang    = "en")
```

The heatmap reveals that most studies addressing chemical agents (Block B)
also address the AM technology context (Block F), while safety risks
(Block A) are largely absent from the literature.

---

# Step 5: Preventive intelligence indicators

## Abstract Sufficiency Score (ASS)

The ASS is a cumulative 0-5 index of preventive informativeness:

```{r ass}
mx  <- orm_ass(result$mx, verbose = TRUE)
orm_ass_plot(mx, lang = "en")
```

**Distribution for the metal AM corpus (N=114):**

| Level | N | % | Meaning |
|-------|---|---|---------|
| 0 | 28 | 24.6% | Non-informative for OHS |
| 1 | 12 | 10.5% | Mentions hazard, no workplace context |
| 2 | 27 | 23.7% | Occupational context present |
| 3 | 24 | 21.1% | Exposure measurement reported |
| 4 | 20 | 17.5% | Worker exposure with result |
| 5 | 3 | 2.6% | Complete: exposure + population + method + prevention |

**Mean ASS = 2.04/5**. Only 3 abstracts (2.6%) contain complete
preventive information. Nearly one quarter (24.6%) contain no
OHS-useful information at all. This quantifies the practical gap
that practitioners face when consulting the primary literature.

## Bridge article detection

```{r bridge}
mx <- orm_bridge(result$mx, verbose = TRUE)
```

**Bridge classification:**

| Type | N | % |
|------|---|---|
| Strong bridge (score 4-5) | 22 | 19.3% |
| Partial bridge (score 3) | 4 | 3.5% |
| Technical study | 88 | 77.2% |

**77.2% of studies are purely technical** -- they do not simultaneously
address workers, exposure measurement, AND preventive recommendations.
Only 22 studies (19.3%) qualify as strong bridges between technical
science and applied OHS prevention.

## Priority reading ranking

```{r ranking}
ranking <- orm_ranking(result$mx, top_n = 10, lang = "en")
print(ranking)
```

The top-ranked article (*Additive Manufacturing for Occupational Hygiene*,
Stefaniak et al. 2021) achieves a perfect bridge score of 5/5 and ASS
of 5/5, with a combined priority score of 21. This is the single most
valuable article for a practitioner entering this field.

---

# Step 6: Practitioner outputs

## Risk sheet

The `orm_risk_sheet()` function generates a structured, actionable risk
sheet for OHS practitioners. It is regulation-neutral (applicable
globally) and includes the traffic-light priority classification:

```{r risk_sheet}
orm_risk_sheet(result,
  topic           = "Metal additive manufacturing",
  search_strategy = "Systematic search in WoS and Scopus (2015-2026).",
  out_dir         = "outputs/",
  lang            = "en"
)
```

The risk sheet:

- Separates **context categories** (RCS > 15, dominant topic) from
  actual **risk categories**
- Classifies each risk as **RED** (critical gap), **AMBER** (attention),
  **GREEN** (reasonable coverage), or **GREY** (insufficient evidence)
- Includes a full **methodological section** with WRDI definition,
  extraction method, deduplication statistics, and limitations

## Guided extraction matrix

For systematic reviewers, `orm_extraction_matrix()` generates a
pre-filled CSV template ready for full-text review:

```{r extraction}
orm_extraction_matrix(result$mx, result,
  top_n   = 20,
  out_dir = "outputs/",
  lang    = "en"
)
```

The matrix pre-fills bibliographic data, ORISMA scores, auto-detected
technology, and risk categories. Reviewers only need to complete the
PDF-dependent fields (study design, population, exposure level, main
result, preventive recommendations).

---

# Step 7: Validation

ORISMA's automatic classification should be validated against manual
review before publication. `orm_validate()` supports this:

```{r validate}
# Step 1: Generate validation sample
val_path <- orm_validate(result$mx,
  n_sample = 30,
  out_dir  = "outputs/validation/",
  lang     = "en"
)

# Step 2: Open the CSV, fill in manual_* columns (0 or 1)
# Step 3: Compute Cohen's Kappa
kappa_results <- orm_validate(result$mx,
  validation_file = val_path,
  out_dir         = "outputs/validation/",
  lang            = "en"
)
print(kappa_results)
```

A Kappa >= 0.7 is considered acceptable for publication in high-impact
OHS journals.

---

# Summary of results

The ORISMA analysis of 114 unique studies on occupational health in
metal additive manufacturing reveals:

1. **High technical-preventive disconnection** (WRDI = 0.61): 60.5% of
   studies lack direct worker exposure data. The evidence base is
   predominantly technical.

2. **Extreme category saturation**: Chemical agents (RCS = 13.0) and
   the AM technology context (RCS = 24.1) dominate the literature.
   Safety, ergonomic, and psychosocial risks are virtually absent.

3. **Critical material gap**: Titanium (MGP = 45.3) is the most
   hazardous and least-studied material. A priority target for future
   research.

4. **Low abstract informativeness** (mean ASS = 2.04/5): Most abstracts
   describe technical processes without conveying worker exposure data.
   Only 2.6% of abstracts are fully informative for OHS practitioners.

5. **Few bridge articles** (22/114, 19.3%): Only one in five studies
   connects technical characterisation with real worker data and
   preventive recommendations.

These findings demonstrate both the utility of ORISMA for rapid
evidence mapping and the significant methodological gap in the
occupational health literature on metal AM.

---

# Session information

```{r session}
sessionInfo()
```

---

# References

Aguilar-Elena, R. & Delgado-Garcia, A. (2025). *Mapping the Safety
Landscape of Emerging Technologies: A Bibliometric Analysis of
Occupational Risks in Metal Additive Manufacturing*. [Under review]

International Organization for Standardization. (2018). *ISO 45001:2018
Occupational health and safety management systems -- Requirements with
guidance for use*. ISO.

National Institute for Safety and Health at Work (INSST). (2023).
*Clasificacion de los accidentes de trabajo por forma/contacto*.
Ministerio de Trabajo y Economia Social, Spain.

National Institute for Occupational Safety and Health (NIOSH). (2023).
*NIOSH Pocket Guide to Chemical Hazards*. CDC/NIOSH.

European Agency for Safety and Health at Work (EU-OSHA). (2023).
*Foresight on new and emerging occupational safety and health risks*.
EU-OSHA.
